---
title: 'Introduction to R: Session 1'
date: "`r format.Date(Sys.Date(), format = '%d %B %Y')`"
author:
  - Alexis Robert
  - Sophie Meakin
  - Carmen Tamayo
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 2
    number-sections: true
    df-print: paged
execute:
  eval: false
  warning: false
  message: false
---

```{r setup, include = FALSE, eval = TRUE}
# Install pacman if not already installed
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman", repos = "https://cloud.r-project.org")
}

# Install and load required packages
pacman::p_load(emoji)
```

# R basics

One of the most important uses of any statistical software, including R, is to analyse data that you may have collected yourself or that have been given to you. In this section, We will introduce how to use R to access data that you already have on your computer, and how to store the results of your analysis or save an updated version of the data, on your computer.

## R Studio

Your R studio session will have four main panels:

1.  **Script** (top-left) - this is where you can write, save and run longer bits of code.
2.  **Environment** (top-right) - this is where you can see objects that you have created in your current R session.
3.  **Console** (bottom-left) - this is where you can write and run very short bits of code.
4.  **Files, plots, etc** (bottom-right) - this is where you can see your files, or where plots will appear.

## Basic R expressions

This section introduces the commands to do simple operations in R, such as arithmetic, or how to make objects.

### Arithmetic operators

These operators are used to carry out mathematical operations, like addition or multiplication:

-   `+`: addition
-   `-`: subtraction
-   `*`: multiplication
-   `/`: division
-   `^`: exponent

Examples:

```{r, eval = FALSE}

46 + 3
13 / 8
2^5

```

`r emoji::emoji("computer")` Try using some of these operators in your R Studio console (the bottom-left panel).

### The assignment operator

`r emoji::emoji("star")` The assignment operator in R is `<-` and is used to assign values to named objects.

`r emoji::emoji("star")` If you run every line in the console (bottom left panel), you will not easily see the previous lines of code, that's why we use the script (top left panel).

-   To execute a line of code from your script: place your cursor on the line you want to run and click on `run` (top right of the script panel) OR place your cursor and use the shortcut `CTRL+ENTER` (`CMD+ENTER` on Mac).
-   To run multiple lines of code, select them with your mouse, and use `CTRL+ENTER` (or click on `run`).
-   To run the whole script, use `source` (top right of the script panel), or use the shortcut `CTRL+SHIFT+ENTER` (`CMD+SHIFT+ENTER` on Mac)

```{r, eval = FALSE}

x <- 10 # The object called x is assigned the value 10
x # Print the object x

y <- 10 + 20 # The object called y is assigned the value 10 + 20
y # Print the object y

z <- "hello" # The object called z is assigned the text "hello"
z # Print the object z

```

`r emoji::emoji("computer")` Look at your Environment panel (top-right) of your R Studio session, then run the above code in your console. Look in the environment panel again: what has changed?

Note that the names of objects are case sensitive (i.e. `Y` and `y` are distinct names) and must not start with numbers or contain spaces. The best approach is to use meaningful object names and stick to using all lower case; if you need to use spaces, then you can use an underscore (`_`).

-   Examples of good object names: `raw_data`, `clean_data`, `country_names`.
-   Examples of bad object names: `raw data`, `CLEAN_data`, `dasyg`, `1x`.

`r emoji::emoji("question")` Discuss *why* each of the examples of bad object names are bad.

`r emoji::emoji("computer")` Try making some of your own objects, e.g. assign the value `"2025"` to the object `year`.

### Relational operators

Relational operators are used to compare different values:

-   `<` less than
-   `>` greater than
-   `<=` less than or equal to
-   `>=` greater than or equal to
-   `==` equal to
-   `!=` not equal to

Expressions using relational operators will return `TRUE` or `FALSE`.

`r emoji::emoji("computer")` Copy the below code into the console of your R Studio terminal.

```{r, eval = FALSE}
x <- 3 # Assign the value 3 to the object called x
y <- 11 # Assign the value 11 to the object called y
```

Examples of using relational operators:

```{r, eval = FALSE}

# Test: is x less than 4?
x < 4

# Test: is y greater than or equal to x
y >= x

# Test: is x equal to 5?
x == 5

# Test: is y not equal to 22
y != 22

```

`r emoji::emoji("computer")` Try using some of these operators in your R Studio console, e.g. to test if `2^11` is bigger than `3*729`.

### The `c()` function

The `c()` function is used to combine multiple values into a *vector* or a *list*. For example, multiple numeric values can be combined into a vector:

```{r, eval = FALSE}

c(3, 1, 10.5)

```

Multiple text strings can also be combined:

```{r, eval = FALSE}

c("Welcome", "to", "R", "!")

```

A vector can be assigned to an object, elements of the vector are accessed with brackets `[]`:

```{r, eval = FALSE}

a <- c(5, 4, 10)
a[1]
a[2] + 3

```

`r emoji::emoji("question")` What is the value of the third element of `a`?

`r emoji::emoji("computer")` Try running `a + 3`, what results do you get?

If both numbers and text are combined with the `c()` function, then all the numeric values are transformed into text:

```{r, eval = FALSE}

b <- c("AFRO", 4)
b
```

`r emoji::emoji("question")` When running `b`, how do you see that the second element was transformed into text?

`r emoji::emoji("question")` What happens if you try adding 2 to the second element of b? Why?

## Inspecting data frames

Datasets are usually stored as data frames, which are objects that can contain several columns (i.e. features) of different types. In this section, we will import a dataframe, and introduce basic functions to inspect the data contained in the dataset.

There are several ways to import data into R, in this session we will import a dataset from a local file. In future sessions, we will explore how to import datasets from online sources or packages.

For this example, we will use a cleaned case linelist dataset that contains epidemiological data from an outbreak. First, we need to load the data using the `readRDS()` function, which reads R data files.

```{r, eval = FALSE}
# Load the linelist data
linelist <- readRDS("data/linelist.rds")
```

The linelist contains case data with multiple variables. We want to only focus on a few columns for this exercise, so we're going to select these columns. You can check the column names of a dataframe by using the function `colnames()`:

```{r, eval = FALSE}
colnames(linelist)
```

We only want to select the case ID, administrative region, age, weight, and gender. Therefore, we create a new object, `data_cases`, which contains the columns of interest. We use the brackets to indicate the columns we are interested in.

```{r, eval = FALSE}
data_cases <- linelist[, c("case_id", "adm1", "age_years", "wt_kg", "gender")]
```

`data_cases` contains all the rows included in linelist, but only the columns case_id, adm1, age_years, wt_kg, and gender. As previously discussed, we should avoid using upper case characters, spaces, brackets, and parentheses when naming objects, since that leads to errors when we try to access these objects. The column names in our dataset already follow good naming conventions, so we don't need to rename them.

Now we have imported our data, we can look at it in one of two ways:

1.  `r emoji::emoji("computer")` Click on the `data_cases` object in the 'Environment' panel (top-right). A new window will open in front of your script pane showing the data. Have a look!
2.  `r emoji::emoji("computer")` Run the command `data_cases` in your Rstudio console. The data will print in the console. Alternatively, the command `head(data_cases)` will print the top 6 lines only:

```{r, eval = FALSE}
head(data_cases)
```

Notice that there are `NA` values in some columns. These indicate missing values, where no data is available.

You can also get the number of columns and number of rows in the data set with `ncol()` and `nrow()`, respectively:

```{r, eval = FALSE}
ncol(data_cases)
```

`r emoji::emoji("computer")` Use `nrow()` to find the number of the rows in the data `data_cases`.

Sometimes we might just be interested in the values of a single column. To extract the values in a single column, we use the `$` operator, e.g. `data_cases$case_id` will return all the values in the column called `case_id`.

`r emoji::emoji("computer")` Run the commands `data_cases$case_id` and `data_cases$wt_kg` in your R console to see what outputs you get (hint: you can use `head(data_cases$case_id)` to only see the first few values).

### Data types

Objects in R are assigned a data type, which tells R how to handle the object.

The simplest data types that you will commonly come across are:

-   *character* i.e. text, e.g. `"f"` or `"London"`
-   *numeric* and *integer*, e.g. `2L`, `1:10`, `2.1`, `pi`
-   *date* e.g. `Sys.Date()`
-   *logical* i.e. `TRUE` or `FALSE`

To check what data type an object is, use the function `class()`.

```{r, eval = FALSE}
class(data_cases$case_id)

class(data_cases$wt_kg)
```

`r emoji::emoji("question")` What data type is the column "age_years" in the `data_cases` data? `r emoji::emoji("question")` What data type is the column "gender" in the `data_cases` data?

`r emoji::emoji("computer")` Use the function `table` to compute the number of cases classified as "m" and "f":

```{r, eval = FALSE}
table(data_cases$gender)
```

`r emoji::emoji("computer")` Combine the functions `prop.table` and `table` to compute the proportion of cases classified as "m" and "f":

```{r, eval = FALSE}
prop.table(table(data_cases$gender))
```

More complex data types can be used to combine multiple bits of data. For instance, objects that contain several elements are called data structures. There are three main types of data structure: - *vectors*: containing one variable per element (created by the function `c()`, like when we created `a` and `b` in the previous section). - *matrices*: containing several variables per element (all variables have the same type). - *data frames*: containing several variables per element (variables can have different types).

`data_cases` contains several variables, of different types (some are characters, others are numeric), it is therefore a data frame. Columns of data frames are vectors:

`r emoji::emoji("question")` What would happen if `data_cases` was a matrix? (hint: try running `head(as.matrix(data_cases))`).

Other data types, such as *factor* and *list*, can be used and are not covered in this practical.

## Basic operations with numeric data

With numeric data, we can calculate different summary statistics, such as the maximum and minimum values, the mean value,and the median value.

### Mean and median

We can calculate the mean value with the function `mean()`.

`r emoji::emoji("computer")` Try running the below command to calculate the mean weight:

```{r, eval = FALSE}
# Calculate the mean weight
mean(data_cases$wt_kg)
```

This will return `NA` - not very helpful! This happens because there are missing values (`NA` values) in `wt_kg` - you can see these by inspecting your data again. If there are any missing values in your data, `mean()` (and many other functions) will simply return `NA`.

To remove these missing (`NA`) values before calculating the mean, we include the argument `na.rm = TRUE`:

```{r, eval = FALSE}
# Calculate the mean weight
mean(data_cases$wt_kg, na.rm = TRUE)
```

`r emoji::emoji("question")` What happens if you change this so that `na.rm = FALSE`?

By default, the value of the argument `na.rm` (standing for NA remove) in `mean()` is set to `FALSE`, so running `mean(data_cases$wt_kg, na.rm = FALSE)` is equivalent to `mean(data_cases$wt_kg)`.

Functions can have many arguments to tackle specific situations (e.g. how to deal with unreported entries). In order to read the description of the expected arguments and outputs of a function, you can use the operator `?`: e.g. `?mean`. It will open a helpfile, which will contain the list of all compatible arguments, and their default value (if any).

`r emoji::emoji("question")` What arguments are compatible with the function `mean` ?

We can calculate other statistics, too. We can calculate the median value with the function `median()`:

```{r, eval = FALSE}
# Calculate the median weight
median(data_cases$wt_kg, na.rm = TRUE)
```

Note that we also have to include the `na.rm = TRUE` argument so that missing (`NA`) values are removed before the median is calculated.

`r emoji::emoji("computer")` Calculate the mean and median age of the cases in `data_cases`. You can remind yourself of the other column names by looking at the data, or checking with the function we introduced earlier.

### Minimum and maximum

We can calculate the minimum and maximum values with the functions `min()` and `max()`, respectively:

```{r, eval = FALSE}
# Calculate the minimum weight
min(data_cases$wt_kg, na.rm = TRUE)

# Calculate the maximum weight
max(data_cases$wt_kg, na.rm = TRUE)
```

`r emoji::emoji("computer")` Calculate the minimum and maximum age for the cases in `data_cases`.

### Standard deviation and inter-quartile range

We can calculate the standard deviation with `sd()` and the inter-quartile range (IQR) with `IQR()`:

```{r, eval = FALSE}
# Calculate the standard deviation of weight
sd(data_cases$wt_kg, na.rm = TRUE)

# Calculate the IQR of weight
IQR(data_cases$wt_kg, na.rm = TRUE)
```

`r emoji::emoji("computer")` Calculate the standard deviation and inter-quartile range for one of the other numeric columns in the `data_cases` data.

A quick way to generate summary statistics for all columns in a data frame is to use the function `summary`:

```{r, eval = FALSE}
# Show the summary of each column in data_cases
summary(data_cases)
```

# R functions

An R function contains a set of commands grouped together in order to perform a specific task. For example, the function `mean()` calculates the mean of some values. To do so, it first adds the values together, and divides by the number of values.

Arguments are the bits of information that you have to give to a function in order for it to work. The function `mean()` has one argument, which tells R the vector of numbers it should use to compute the mean: `mean(x)` therefore computes the mean value of the vector `x`. It also has an optional argument: `na.rm`, which is set to FALSE unless specified by the user. This means that the function `mean` will not remove the unreported (NA) values by default.

**Extra:** We can write our own functions, if we need them! In the example below, the function `sum_pow` takes two arguments (called `a` and `b`), computes their squared value., and returns the sum of the squared values.

```{r, eval = FALSE}
sum_pow <- function(a, b) {
  ## Compute a squared
  a_squared <- a^2
  ## Compute b squared
  b_squared <- b^2

  ## Compute the sum of the squared values
  result <- a_squared + b_squared

  return(result)
}

sum_pow(3, 7)
sum_pow(10, 17)
```

`r emoji::emoji("computer")` Copy and paste the code above into your RStudio script and run it.

`r emoji::emoji("question")` Optional: Can you create a function that takes 3 arguments (for instance `a`, `b`, and `c`) and returns the sum of their squared values?

# R packages

R packages are extensions of the set of base functions contained in R by default. They are mostly developed by the community, and extend what R can do. Packages typically contain functions, but can also contain datasets. Some packages contain a few functions that facilitate specific tasks that would be tedious using only base R, while others create new possibilities (like mixing R and other coding languages, or introducing new operators).

We use the command `install_packages` to download a package on the computer. At that stage, the package is not yet imported in the R session, it just becomes accessible on your computer. To import the package on the current session, we use the function `library`. In this session, we will use the package **rio** to import datasets, let's make sure you are all able to install it.

`r emoji::emoji("computer")` Copy and paste the code below.

```{r, eval = FALSE}
# Install the package rio
install.packages("rio")

# Import rio into the current session
library(rio)
```

Packages have to be imported (using `library`) every time you open a new R session, but they only need to be installed once. 

If you're not certain what a package does or contains you can view the package documentation by using `?`, or `help()`.

`r emoji::emoji("computer")` Check the description of the **rio** package by running `?rio` or `help(rio)`.

# Working directory

The working directory is the root folder used by R, and will be the default location where R will start to look for data or saves files. You can find out your current working directory with the function `getwd()`:

```{r, eval = FALSE}

getwd()

```

```{r, eval = FALSE}

## [1] "C:/Users/eidearob/Documents"

```

`r emoji::emoji("question")` Use `getwd()` to find out your working directory. How is it different from the one above? Check whether you have the same working directory as other students.

You would want the working directory to be in the folder `sitrep`, so R can easily access all the scripts, data, and output contained in the folder. If it was in the `script` folder, then it would not be able to access the `data` folder. To change the working directory, you can use `setwd()`, a function that takes an argument called `dir`, the path to the new working directory.

If you are not sure where path to the `sitrep` folder is on your computer, you can use the tab `session` (at the top of your screen), then click on `Set Working Directory` and `Choose Directory`. This way, you will be able to browse your computer and locate the folder you created previously.

If you are still not certain where your `sitrep` folder is located, you can use the tab `session` (at the top of your screen), then click on `Set Working Directory` and `To Source File Location`. In your script, you will see the command run by your computer leading to the `script` folder (where you saved your current script in the previous section). Just remove `/script` from the command, and your working directory will be in `sitrep`.

`r emoji::emoji("question")` Set your working directory to `sitrep`. Use `getwd()` to check you're in the right working directory. Is your working directory now the same as other students?

# Importing and exporting data

## Importing data with **rio**

We can import (i.e. 'read in') and export (i.e. save) data in common formats (e.g. .csv, .xlsx, .txt) with the **rio** package. In this section, we are going to import some data on life expectancy at birth and health expenditure per capita in USD for countries in 2015. This data is from Our World in Data: [ourworldindata.org/life-expectancy-vs-health-expenditure](https://ourworldindata.org/grapher/life-expectancy-vs-health-expenditure).

`r emoji::emoji("star")` We import a data file (e.g. .csv) using the `import()` function from the **rio** library.

`r emoji::emoji("computer")` Download the file `life_expectancy.csv` from Moodle, and place it in your Data folder.

`r emoji::emoji("computer")` Copy and paste the code below. This code imports the data `life_expectancy.csv` from the folder `data`, and assigns it to an object called `life_expectancy`:

```{r, eval = FALSE}

life_expectancy <- import(file = "data/life_expectancy.csv")

```

Let's look more closely at how `import()` works.

### The `import()` function

The `import()` function has one main argument: `file`. This argument is where we tell R where to find the file we want to import.

If you use R a lot, you might see people using other functions or libraries, such as the function `read.csv()` for .csv files only, or the library [`readxl`](https://readxl.tidyverse.org/) for .xls or .xlsx files only. But one of the great things about `import()` is that you can use it for all common data formats e.g. `.csv`, `.txt`, `.xlsx`, so you only have to remember that!

## Exporting data

`r emoji::emoji("star")` We export data using the `export()` function from the **rio** library.

The argument `x` specified the name of the data (e.g. the data frame) that we want to export. The argument `file` tells R where to save the data and what to call it.

The code below exports the data frame `life_expectancy` and saves it as a .csv file called "test_data_export.csv" in the data folder:

```{r, eval = FALSE}
export(x = life_expectancy, file = "data/test_data_export.csv")
```

# Investigating `life_expectancy`

Now that we have imported a new data file, let's explore what data it contains. In the section abovel, we used several functions to describe the **linelist** dataset:

-   `head()` shows the first 6 rows of the dataframe, use the argument `n` to show more than 6 rows.

-   `colnames()` shows the column names of the dataframe.

-   `ncol()` and `nrow()` respectively computes the number of rows (i.e. the number of observations) and the number of columns (i.e. the number of variables) in the dataframe.

-   `class()` returns the class of the object.

-   `mean()` and `median()` computes the mean and median of a vector.

-   `min()` and `max()` computes the minimum and maximum value of a vector.

-   `sd()` and `IQR()` computes the standard deviation and the interquartile range of a numeric vector.

-   `summary()` returns the minimum, first quartile, median, mean, third quartile, and maximum value of a vector.

-   `table()` returns the number of entries for each value of a vector.

Now, we're going to use these functions to describe the `life_expectancy` dataset. First, let's have a look at what the data looks like by using `head()`

```{r, eval = FALSE}
head(life_expectancy)
```

This shows us that there are five columns in this dataset, and each row describes one country.

`r emoji::emoji("question")` What values are shown in the column `health_expenditure_usd`? What does that mean?

Now, let's describe the data contained in `life_expectancy`:

`r emoji::emoji("question")` Using `nrow()`, how many observations are there in the dataset?

`r emoji::emoji("question")` Using `class()`, which columns in the dataset are numeric?

`r emoji::emoji("question")` Using `table()`, what are the regions (column `region`) contained in the dataset? How many entries are there by region? (hint: we are only interested in the column `region`, use the operator `$` to select that column from `life_expectancy`). Entering `table(life_expectancy)` would create a table across the whole dataframe, which is not what we are interested in.

As you observed when we looked at the first few rows, there are unreported entries in the columns `life_expectancy_years` and `health_expenditure_usd`. However, we'd like to know how many entries are missing in each column. That will tell us whether this dataset can provide a good description of these values, or if there are too many missing entries to make any analysis on this data.

The function `is.na()` returns `TRUE` for each `NA` value in the argument. We can use `table()` and `is.na` to know the number of `NA` values in `life_expectancy_years`.

`r emoji::emoji("computer")` Run the code below to compute the number of `NA` values in `life_expectancy_years`. You should get 17 NA values out of 217 rows

```{r, eval = FALSE}
table(is.na(life_expectancy$life_expectancy_years))
```

`r emoji::emoji("question")` Compute the number of `NA` values in the column `health_expenditure_usd`. How does this compare to the column `life_expectancy_years`? Given the number of unreported entries in `health_expenditure_usd`, can we use this dataframe to give an accurate description of the distribution of health expenditure in the world?

`r emoji::emoji("question")` Are there any missing entries in the other columns of the dataframe?

Let's first describe the column `life_expectancy_years`:

`r emoji::emoji("question")` Using `min()`, `max()`, `IQR()`, `mean()`, and `median()`, describe the column `life_expectancy_years` of the dataframe. What is the average life expectancy across all countries? Describe the dispersion of the values.

Describing the entire dataframe is informative, but it may be more interesting to describe the regions separately, and see the countries that have extreme values of `life_expectancy` or `health_expenditure`.

`r emoji::emoji("computer")` In base R, we can select certain rows using `==`, for instance, Use the following command to select only the rows describing countries from the region Europe and Central Asia:

```{r, eval = FALSE}
## Inside the brackets[]:
## - conditions before the comma select rows,
## - conditions after the comma select columns
## Here, we select the rows where the region is Europe and Central Asia, and select all the columns
life_expectancy[life_expectancy$region == "Europe and Central Asia", ]
```

This shows us a subset of `life_expectancy`.

`r emoji::emoji("computer")` Use the following command to compute the average life expectancy in Europe and Central Asia

```{r, eval = FALSE}
mean(life_expectancy[life_expectancy$region == "Europe and Central Asia", ]$life_expectancy_years,
  na.rm = TRUE
)
```

`r emoji::emoji("question")` What is the mean life expectancy in East Asia and Pacific according to our dataframe? What is their average health expenditure?

According to our dataframe, the average per capita health expenditure in East Asia and Pacific is about 3,759USD. However, we saw that lots of entries in `health_expenditure_usd` were missing.

`r emoji::emoji("computer")` Use the code below to compute the number of missing entries of `health_expenditure_usd` in East Asia and Pacific. Is there enough reported entries to tell us something about health expenditure across countries in East Asia and Pacific?

```{r, eval = FALSE}
table(is.na(life_expectancy[life_expectancy$region == "East Asia and Pacific", ]$health_expenditure_usd))
```

Finally, we want to look at the extreme values in our dataframe: what countries have the longest average values of life expectancy? What countries report shorter values? To do so, we can use the function `order()` to order the rows. `order()` returns the position of each value in the vector, by default using an increasing order (the smallest value is assigned the value 1, the next smallest is 2 and so on).

`r emoji::emoji("computer")`: Run the following command to order the `life_expectancy` dataframe by life expectancy.

```{r, eval = FALSE}
life_expectancy[order(life_expectancy$life_expectancy_years), ]
```

If we are interested in countries with the highest values of life expectancy, it may be more convenient to order the dataframe in decreasing order (i.e. highest values are shown first). To do so, we can use the argument `decreasing` from the function `order()`.

```{r, eval = FALSE}
life_expectancy[order(life_expectancy$life_expectancy_years, decreasing = TRUE), ]
```

We can now use the function `head()` to only see the rows that have the highest life expectancy.

`r emoji::emoji("question")` What are the highest values of life expectancy in the dataframe?

```{r, eval = FALSE}
head(life_expectancy[order(life_expectancy$life_expectancy_years, decreasing = TRUE), ])
```

Finally, let's find the extreme values per region. We can combine the selection process introduced earlier (using `==`) and the `order()` function. To avoid creating long lines of code that may be hard to understand for collaborators (or for yourself if you re-open your script in 6 months), we can use two different commands: First we create a new dataframe containing only the entries in East Asia and Pacific, then we order this new dataframe by `life_expectancy_years`.

```{r, eval = FALSE}
# Create a new dataframe containing only countries from East Asia and Pacific
life_exp_asia_pacific <- life_expectancy[life_expectancy$region == "East Asia and Pacific", ]

# Order the newly created dataframe by life expectancy
life_exp_asia_pacific[order(life_exp_asia_pacific$life_expectancy_years, decreasing = TRUE), ]
```

`r emoji::emoji("question")` What countries report the highest life expectancy in the region `Europe and Central Asia`? And in the region Latin America and Caribbean?

# Investigating `measles_per_year`

In the last section of this practical, we will describe the dataset `measles_per_year`. This dataset contains the number of measles cases per year per WHO region, and was downloaded from [ourworldindata.org](https://ourworldindata.org/grapher/reported-cases-of-measles).

`r emoji::emoji("computer")` Download the file `measles_per_year.csv` from Moodle, and place it in your Data folder. Import it into your R session using the function `import`:

```{r, eval = FALSE}

measles_cases <- import(file = "data/measles_per_year.csv")

```

Answer the following questions using the functions presented in this practical and in Intro to R 1:

`r emoji::emoji("question")`: How many observations (rows) are there in the dataset?

`r emoji::emoji("question")`: What are the column names of the dataset?

The column names do not follow the best coding practice, let's change them to something shorter and more consistent, to avoid future errors:

```{r, eval = FALSE}
colnames(measles_cases) <- c("region", "year", "cases")
```

`r emoji::emoji("question")`: What is the class of each column in the dataset?

`r emoji::emoji("question")`: Are there any `NA` entries in the `cases` column? (hint: use the functions `table()` and `is.na()`)

`r emoji::emoji("question")`: Describe the column `cases`: what is the median number of annual cases in the dataset? What is the maximum / minimum values? What is the IQR?

The dataset contains the number of cases per WHO region, let's look at how they changed through time:

`r emoji::emoji("question")`: Describe the column `cases` in the `region` `Europe`: What was the minimum and maximum number of annual cases in Europe?

`r emoji::emoji("question")`: Observe the number of cases per year in Europe: using `measles_cases[measles_cases$region == "Europe", ]`, did the number of measles cases decrease between 1990 and 2019? Did the number of cases decrease between 2007 and 2019?

`r emoji::emoji("question")`: Was the dynamic similar in `Africa`: did the number of cases decrease between 1980 and 2020?

`r emoji::emoji("question")`: Same question if the region is `Americas`: How does the number of cases in `Americas` compare to the number of cases in `Europe`?

Check to see whether the other three regions (`South-east Asia`, `Western Pacific`, and `Eastern Mediterranean`) have similar dynamics

Let's summarise our findings, we can deduce various characteristics of measles dynamics since 1990:

-   Across all regions, the number of measles cases per year greatly decrease between 1980 and 2000.

-   From 2000 onwards, the dynamics changed between regions: the decrease continued in certain regions (e.g Europe), but had already reached a very low level in the Americas

-   Since 2010, we observe cyclical increase and decreases in the number of annual cases in each region.

-   There are great discrepancies between regions: in some regions there is low incidence despite recurring outbreaks (Americas), while others still report a large number of cases every year.

-   Some regions reported very few cases in 2021, which could be linked to the COVID-19 pandemic.
