---
title: "Task 5: Getting epidemiological parameters"
format:
  html:
    code-fold: true
execute:
  warning: false
---

## Overview

This task focuses on obtaining epidemiological parameters from your data or the literature. You will learn how to extract, estimate, and use key epidemiological parameters in your analyses.

## Learning Objectives

- Extract epidemiological parameters from data
- Search and retrieve parameters from literature
- Estimate parameters using available data
- Incorporate parameters into analytical models

## Tutorial

[Access the tutorial here](https://epiverse-trace.github.io/tutorials-middle/delays-access.html)

## Example Report

<a href="data/solutions/task5_solution.qmd" download class="btn btn-primary" target="_blank">Download solution file</a>

### Ebola serial interval
The serial interval for Ebola was extracted from the `epiparameter` package, identifying the following source:

```{r}
#| results: asis

# INSTALL AND LOAD PACKAGES

# Load required packages
pacman::p_load(epiparameter, tidyverse, EpiNow2, reactable)

# Creating epiparameter objects by accessing the {epiparameter} library
ebola_serial_int <- epiparameter::epiparameter_db(
  disease = "ebola",
  epi_name = "serial",
  subset = sample_size > 10,
  author = "WHO Ebola Response Team",
  single_epiparameter = TRUE
)

ebola_serial_int$citation
```

```{r}
# Extract summary statistics and parameters
pars <- epiparameter::get_parameters(ebola_serial_int) # distribution parameters
serial_mean <- ebola_serial_int$summary_stats$mean
serial_sd <- ebola_serial_int$summary_stats$sd

# Calculate 95th percentile of serial interval
serial_interval_95th <- qgamma(
  0.95,
  shape = pars["shape"],
  scale = pars["scale"]
)
```

This estimate has a mean of `r round(serial_mean, 1)` days and standard deviation of `r round(serial_sd, 1)` days, meaning that 95% of cases have a serial interval of less than `r round(serial_interval_95th, 0)` days. The probability density function of the serial interval is given below:

```{r}
plot(ebola_serial_int) # as a plot of the PDF
```
