---
title: "Task 7: Nowcasting epidemiological data"
format:
  html:
    code-fold: true
execute:
  warning: false
---

## Overview

This task walks you through nowcasting epidemiological data. 
You will learn to correct for epidemiological and reporting delays so you can 
estimate current case counts even when the data are incomplete. 
Along the way, you will learn how to work with the delay distributions provided 
by {epiparameter}—discretising continuous distributions, turning parameters 
into {EpiNow2} inputs, and using those distribution functions to guide decisions
like setting appropriate quarantine lengths.

## Learning Objectives

- Understand reporting delays and their impact
- Apply nowcasting methods to epidemiological data
- Adjust for underreporting and delays
- Generate real-time estimates of current incidence

## Tutorial

[Access the tutorial here](https://epiverse-trace.github.io/tutorials-middle/delays-functions.html)

## Example Report

### Visualising available epiparameter distributions

 We first visualise the available distributions in {epiparameter} that we can 
 use to adjust for delays during an Ebola outbreak.

```{r}
#| echo: true
#| eval: true
#| message: false
#| warning: false

# INSTALL AND LOAD PACKAGES

if (!"tidyverse" %in% rownames(installed.packages())) {
  install.packages("tidyverse")
}

if (!"EpiNow2" %in% rownames(installed.packages())) {
  install.packages("EpiNow2")
}

if (!"incidence2" %in% rownames(installed.packages())) {
  install.packages("incidence2")
}

if (!"kableExtra" %in% rownames(installed.packages())) {
  install.packages("kableExtra")
}

# Load required packages
library(tidyverse)
library(EpiNow2)
library(incidence2)
library(kableExtra)

# List distributions
epiparameter::epiparameter_db(disease = "ebola") %>%
  epiparameter::parameter_tbl() %>%
  knitr::kable(format = "html", caption = "Ebola delay distributions") %>%
  kableExtra::kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

# Extracting relevant parameters and adjusting for delays

We imported linelist data for the Ebola outbreak, and converted it to incidence.
Then extracted parameter distributions for the serial interval and incubation
period for Ebola, chosen from the table above.
We also adjust for a reporting delay of 2 days from onset to notification,
on average.

```{r}
# LOAD DATA
# Import linelist data and convert to incidence
ebola_incidence <- readRDS("data/linelist.rds") %>%
  incidence2::incidence(date_index = "date_onset",
    count_values_to = "confirm",
    date_names_to = "date",
    complete_dates = TRUE) %>%
  dplyr::select(-count_variable)

# serial interval ---------------------------------------------------------

# subset one distribution for the SI
ebola_serial <- epiparameter::epiparameter_db(
  disease = "ebola",
  epi_name = "serial",
  single_epiparameter = TRUE
)

# adapt SI epiparameter object to EpiNow2
ebola_serial_discrete <- epiparameter::discretise(ebola_serial)

serial_interval_ebola <-
  EpiNow2::Gamma(
    mean = ebola_serial$summary_stats$mean,
    sd = ebola_serial$summary_stats$sd,
    max = quantile(ebola_serial_discrete, p = 0.99)
  )

# incubation period ---------------------------------------------------------

# subset one distribution for ebola incubation period
ebola_incubation <- epiparameter::epiparameter_db(
  disease = "ebola",
  epi_name = "incubation",
  single_epiparameter = TRUE
)

# adapt IP epiparameter object to EpiNow2
ebola_incubation_discrete <- epiparameter::discretise(ebola_incubation)

incubation_period_ebola <-
  EpiNow2::Gamma(
    mean = ebola_incubation$summary_stats$mean,
    sd = ebola_incubation$summary_stats$sd,
    max = quantile(ebola_serial_discrete, p = 0.99)
  )

# reporting delay ---------------------------------------------------------
reporting_delay <- EpiNow2::LogNormal(
  mean = 2,
  sd = 1,
  max = 10
)

# nowcasting with epinow ---------------------------------------------------
withr::local_options(base::list(mc.cores = 4))
epinow_estimates <- EpiNow2::estimate_infections(
  data = ebola_incidence,
  generation_time = EpiNow2::generation_time_opts(serial_interval_ebola),
  delays = EpiNow2::delay_opts(incubation_period_ebola + reporting_delay),
  forecast = NULL,
  stan = EpiNow2::stan_opts(samples = 500, chains = 2) 
)
```

## Additional Resources

::: {.panel-tabset}

### R Handbook
Relevant chapters from the Epidemiologist R Handbook.

### Package Documentation
Package vignettes and documentation for relevant R packages.

:::

