---
title: "Task 7: Nowcasting epidemiological data"
format:
  html:
    code-fold: true
execute:
  warning: false
editor: 
  markdown: 
    wrap: 72
---

## Overview

This task walks you through nowcasting epidemiological data. You will
learn to correct for epidemiological and reporting delays so you can
estimate current case counts even when the data are incomplete. Along
the way, you will learn how to work with the delay distributions
provided by {epiparameter}—discretising continuous distributions,
turning parameters into {EpiNow2} inputs, and using those distribution
functions to guide decisions like setting appropriate quarantine
lengths.

## Learning Objectives

-   Understand reporting delays and their impact
-   Apply nowcasting methods to epidemiological data
-   Adjust for underreporting and delays
-   Generate real-time estimates of current incidence

## Tutorial

[Access the tutorial
here](https://epiverse-trace.github.io/tutorials-middle/delays-functions.html)

## Downloading Nowcast Estimates

**Note:** Running EpiNow2 can be computationally expensive and may take
several minutes to complete. If you want to skip this step, you can
download the pre-computed estimates file and place it in your `outputs/`
folder:

<a href="outputs/nowcast_estimates.rds" download class="btn btn-primary" target="_blank">Download
nowcast estimates</a>

## Example Report

<a href="data/solutions/task7_solution.qmd" download class="btn btn-primary" target="_blank">Download
solution file</a>

### Visualising available epiparameter distributions

We first explore the available distributions in {epiparameter} that we
can use to adjust for delays during an Ebola outbreak.

```{r}
# INSTALL AND LOAD PACKAGES
pacman::p_load(tidyverse, EpiNow2, incidence2, reactable, rio)

# List distributions
epiparameter::epiparameter_db(disease = "ebola") |>
  epiparameter::parameter_tbl() |>
  select(-disease) |>
  reactable::reactable(defaultPageSize = 5)
```

### Extracting relevant parameters and adjusting for delays

We imported linelist data for the Ebola outbreak, and converted it to
incidence. Then extracted parameter distributions for the serial
interval and incubation period for Ebola, chosen from the table above.

```{r}
# LOAD DATA
# Import linelist data and convert to incidence
ebola_incidence <- readRDS("data/linelist.rds") |>
  incidence2::incidence(
    date_index = "date_onset",
    count_values_to = "confirm",
    date_names_to = "date",
    complete_dates = TRUE
  ) |>
  dplyr::select(-count_variable)

# serial interval ---------------------------------------------------------

# subset one distribution for the SI
ebola_serial <- epiparameter::epiparameter_db(
  disease = "ebola",
  epi_name = "serial",
  single_epiparameter = TRUE
)

# adapt SI epiparameter object to EpiNow2
ebola_serial_discrete <- epiparameter::discretise(ebola_serial)

serial_interval_ebola <-
  EpiNow2::Gamma(
    mean = ebola_serial$summary_stats$mean,
    sd = ebola_serial$summary_stats$sd,
    max = quantile(ebola_serial_discrete, p = 0.99)
  )

# incubation period ---------------------------------------------------------

# subset one distribution for ebola incubation period
ebola_incubation <- epiparameter::epiparameter_db(
  disease = "ebola",
  epi_name = "incubation",
  single_epiparameter = TRUE
)

# adapt IP epiparameter object to EpiNow2
ebola_incubation_discrete <- epiparameter::discretise(ebola_incubation)

incubation_period_ebola <-
  EpiNow2::Gamma(
    mean = ebola_incubation$summary_stats$mean,
    sd = ebola_incubation$summary_stats$sd,
    max = quantile(ebola_serial_discrete, p = 0.99)
  )

# reporting delay ---------------------------------------------------------
reporting_delay <- EpiNow2::LogNormal(
  mean = 10,
  sd = 3,
  max = 15
)

# check if the analysis has been run before and run EpiNow2 if not
file_name <- "outputs/nowcast_estimates.rds"
if (file.exists(file_name)) {
  nowcast_estimates <- import(file_name)
} else {
  # nowcasting with epinow ---------------------------------------------------
  withr::local_options(base::list(mc.cores = 4))
  nowcast_estimates <- EpiNow2::estimate_infections(
    data = ebola_incidence,
    generation_time = EpiNow2::generation_time_opts(serial_interval_ebola),
    delays = EpiNow2::delay_opts(incubation_period_ebola + reporting_delay),
    forecast = NULL,
    stan = EpiNow2::stan_opts(samples = 500, chains = 2)
  )
  export(nowcast_estimates, file_name)
}

```

We then adjust for a reporting delay of 10 days from onset to
notification, on average.

```{r}
infection_plot <- plot(nowcast_estimates, type = "infections")

weekly_obs <- nowcast_estimates$observations %>%
  mutate(week= floor_date(date, "week", week_start = 1)) %>%
  group_by(week) %>%
  summarise(
    confirm = mean(confirm, na.rm = TRUE),  # average daily cases that week
    .groups = "drop"
  )

infection_plot$layers <- infection_plot$layers[-1]

infection_plot +
  geom_col(
    data = weekly_obs,
    aes(x = week, y = confirm),
    inherit.aes = FALSE,
    fill = "#9ca3af",
    alpha = 0.4,
    width = 6
  )
```
