---
title: "Task 3: Generating interactive maps"
format:
  html:
    code-fold: true
execute:
  warning: false
  message: false
---

## Overview

This task focuses on creating interactive maps for visualizing spatial aspects of epidemiological data. You will learn how to work with spatial data, create maps using GIS tools in R, and visualize disease patterns geographically.

## Learning Objectives

- Work with spatial data formats (shapefiles, coordinates)
- Create interactive maps using GIS tools in R
- Visualize disease patterns and hotspots geographically
- Perform basic spatial analysis

## Tutorial

[Access the tutorial here](https://www.epirhandbook.com/en/new_pages/gis.html)

## Download GIS Data

For this task, you will need the GIS shapefile data to create the maps. The shapefile contains administrative boundaries for Sierra Leone.

1. Download the GIS data zip file using the button below.
2. **Unzip the file** and extract its contents to your `data/gis/` folder.
3. The extracted files should include `sle_adm3.shp` and other related shapefile components in the `data/gis/` directory.

[Download GIS data (zip)](https://github.com/WHO-Collaboratory/afro-collaborative-analytics/raw/refs/heads/main/qmd/data/gis.zip){.btn .btn-primary target="_blank"}

## Example Report

<a href="data/solutions/task3_solution.qmd" download class="btn btn-primary" target="_blank">Download solution file</a>

### Geographic distribution of cases

```{r}
# INSTALL AND LOAD PACKAGES

# Load required packages
pacman::p_load(
  tidyverse, # basic data manipulation
  sf, # handling shapefiles
  here, # handing file paths
  leaflet, # mapping
  leaflet.extras, # additional leaflet features including heatmaps
  htmltools, # handlers for html objects
  janitor # name cleaning
)


# LOAD DATA

# load shapefiles and keep relevant areas
shp <- sf::read_sf("data/gis/sle_adm3.shp") |>
  # standardize column names
  janitor::clean_names()

# import the cleaned linelist data
linelist <- rio::import("data/linelist.rds")

# convert to aggregate format
agg <- linelist |>
  # count the number of cases per adm3
  summarise(n = n(), .by = adm3) |>
  # add the shape files of the adm3 zones
  inner_join(shp, by = c(adm3 = "adm3_en")) |>
  # convert back to shapefile
  st_as_sf()
```

Total incidence to date was mapped by chiefdom.

```{r}
# define colour palette
pal <- colorNumeric(palette = "YlOrRd", domain = agg$n)

# generate choropleth map (polygons only)
leaflet(agg) |>
  # use CartoDB for background
  addProviderTiles("CartoDB.Positron") |>
  # add adm3 shaded by number of cases
  addPolygons(
    fillColor = ~ pal(n),
    color = "white",
    weight = 1,
    fillOpacity = 0.7,
    # define labels
    label = ~ map(
      paste0("<b>", adm3, "</b><br>", n, " cases"),
      htmltools::HTML
    ),
    labelOptions = labelOptions(direction = "auto"),
    highlightOptions = highlightOptions(
      weight = 3,
      color = "black",
      bringToFront = FALSE
    )
  ) |>
  # add legend
  addLegend(
    pal = pal,
    values = ~n,
    title = "Case count",
    opacity = 0.7,
    position = "bottomright"
  )

```

Spatial clustering of cases was then investigated at a higher spatial resolution using a heatmap.

```{r}
# Prepare linelist data for heatmap
# Filter cases with valid coordinates
linelist_coords <- linelist |>
  filter(!is.na(lon), !is.na(lat))

# Generate heatmap from individual cases
leaflet(linelist_coords) |>
  # use CartoDB for background
  addProviderTiles("CartoDB.Positron") |>
  # add heatmap layer
  addHeatmap(
    lng = ~lon,
    lat = ~lat,
    intensity = 0.5,
    blur = 5,
    max = 0.05,
    radius = 7,
    gradient = c("blue", "cyan", "yellow", "red")
  )

```
