---
title: "Task 9 Solution"
format:
  html
execute:
  echo: false
  warning: false
  message: false
---


### Define population demographics and initial population conditions

We first extract a social contact matrix from {socialmixr} for the United Kingdom. 
A single individual from the youngest age category (under 20 years old) was infected and introduced the disease in a population of 1 million, which were initially fully susceptible to the infection.


```{r}
# INSTALL AND LOAD PACKAGES

# Install epidemics package from GitHub (not available on CRAN)
if (!requireNamespace("epidemics", quietly = TRUE)) {
  # Install remotes if needed
  if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
  }
  # Install epidemics from GitHub
  remotes::install_github("epiverse-trace/epidemics")
}

# Load required packages
pacman::p_load(tidyverse, epidemics, socialmixr)

# Defining social contact matrix
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  survey = polymod,
  countries = "United Kingdom",
  age.limits = c(0, 20, 40),
  symmetric = TRUE
)
# prepare contact matrix
contact_matrix <- t(contact_data$matrix)

# Define population's initial conditions
initial_i <- 1e-6
initial_conditions_inf <- c(
  S = 1 - initial_i, E = 0, I = initial_i, R = 0, V = 0
)
initial_conditions_free <- c(
  S = 1, E = 0, I = 0, R = 0, V = 0
)

# combine the initial conditions
initial_conditions <- rbind(
  initial_conditions_inf, # age group 1
  initial_conditions_free, # age group 2
  initial_conditions_free # age group 3
)

# use contact matrix to assign age group names
rownames(initial_conditions) <- rownames(contact_matrix)

# Create demography vector
demography_vector <- contact_data$demography$population
names(demography_vector) <- rownames(contact_matrix)

# Create population object for {epidemics}
uk_population <- population(
  name = "UK",
  contact_matrix = contact_matrix,
  demography_vector = demography_vector,
  initial_conditions = initial_conditions
)

uk_population
```

### Simulating transmission

We fitted an SEIR to model the transmission of an infectious disease with a basic reproduction number of 1.5, an infectious period of 7 days, and a pre-infectious period of 3 days.

```{r}
# Defining model parameters
# time periods
preinfectious_period <- 3.0
infectious_period <- 7.0

# specify the mean and standard deviation of R0
r_estimate_mean <- 1.5
r_estimate_sd <- 0.05

# Generate 100 R samples
r_samples <- withr::with_seed(
  seed = 1,
  rnorm(
    n = 100, mean = r_estimate_mean, sd = r_estimate_sd
  )
)

# rates
infectiousness_rate <- 1.0 / preinfectious_period
recovery_rate <- 1.0 / infectious_period
beta <- r_samples / infectious_period

# running {epidemic} default model
output <- model_default(
  population = uk_population,
  transmission_rate = beta,
  infectiousness_rate = infectiousness_rate,
  recovery_rate = recovery_rate,
  time_end = 600, increment = 1
)

# Visualising output
output %>%
  mutate(r_value = r_samples) %>%
  unnest(data) %>%
  filter(compartment == "infectious") %>%
  ggplot() +
  geom_line(
    aes(time, value, color = r_value, group = param_set),
    alpha = 3
  ) +
  scale_color_fermenter(
    palette = "RdBu",
    name = "R"
  ) +
  scale_y_continuous(
    labels = scales::comma
  ) +
  facet_grid(
    cols = vars(demography_group)
  ) +
  theme_bw() +
  labs(
    x = "Simulation time (days)",
    y = "Individuals"
  )
```

The figure above shows the total number or cumulative amount of individuals in the infectious compartment at each time, across the 100 simulations run for different basic reproduction number values.
